Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-15 paper-correlations.tex
No connected components!

Cluster function can be calculated with the help of previuously described
procedures.  Firstly, we apply an indicator function to an input $A$:
$A^{(i)} \xleftarrow{I^{(i)}} A$. Then we label clusters in $A^{(i)}$ using a
function that is a part of Julia \code{ImageMorphology.jl} library.  Labeling
algorithm must take into account if the function is calculated with periodic
boundary conditions \cite{evstigneev2023}. For periodic boundary conditions we
implemented our own solution as part of the package, as it is absent in the
public domain. Then $C_2$ function is a sum of $S_2$ functions calculated for
all clusters separately. Algorithmic complexity of this implementation is
$O(MC\log M)$ where $M$ is the number of elements in the input and $C$ is the
number of clusters. The algorithm is then written as:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

@@

Majority of described algorithms contain parallelizable operations such as fast
Fourier transform, connected components labeling and image filtering, hence they
are perfectly suited for execution on GPUs and provide very fast computations
(see execution times below).

\subsection{Directional scanning with line segment}
\label{sec:scan}
All algorithms described above (full correlation maps) are executed on the whole
input image and hence require a lot of RAM memory. The scan approach deals with
memory limitations at the cost of higher execution time (except for $L_2$). With
line-segment scanning we select a list of predefined directions and cut all
one-dimensional slices from the input along those directions. Then for each
direction, as shown in \cref{fig:directions}, we use an appropriate algorithm in
\cref{sec:map} to calculate a correlation function just for that single
slice. For this reason we do not repeat algorithms for $S_2$, $\rho_{ij}$,
$F_{ss}$ and $F_{sv}$.  When all slices are processed we calculate the resulting
directional or averaged CF using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g., image filtering
for edge detection or connected components labeling performed on the whole
array, not on slices.

@@

Functions which calculate full correlation maps are suitable for running on both
CPU and GPU, so we provide execution times for both. The cluster function uses
an algorithm for closest component labeling which has CPU-only implementation
(thus, leaving a room for future improvements). Directional computations are
performed exclusively on CPUs.

@@

While the toolbox of presented package and its API are solid, based on
additional research numerous areas for future improvements of
\code{CorrelationFunctions.jl} are possible:
\begin{enumerate}
\item include novel non-classical correlation functions, e.g., describing
  topology of phases;
\item implementing sub-cube based computations of CFs with a given correlation
  functions with later averaging for larger volume - with such domain
  decomposition one could compute CFs for larger samples with limited GPU
  resources;
\item implementing GPU-based clustering (cluster labeling) and distance-map
  transform to further speed up computations for $C_2$, chord-length and
  pore-size functions;
\item implementing higher order functions, such as 3-point probabilities or even
  3-point cluster and other more involved CFs;
\item increase number of directions for scan approach or implementing
  calculations within a given solid angle.
\end{enumerate}
We perform steps forward in some directions, whilst others still require
additional research to be made to facilitate their implementation within the
package.

>>>>

Cluster function can be calculated with the help of previuously described
procedures.  Firstly, we apply an indicator function to an input $A$:
$A^{(i)} \xleftarrow{I^{(i)}} A$. Then we label clusters in $A^{(i)}$ using a
function \code{label\_components} which is a part of Julia
\code{ImageMorphology.jl} library.  Labeling algorithm must take into account if
the function is calculated with periodic boundary conditions
\cite{evstigneev2023}. For periodic boundary conditions we implemented our own
solution as part of the package, as it is absent in the public domain. Then
$C_2$ function is a sum of $S_2$ functions calculated for all clusters
separately. Algorithmic complexity of this implementation is $O(MC\log M)$ where
$M$ is the number of elements in the input and $C$ is the number of
clusters. The algorithm is then written as:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $Labels \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(Labels)$ \Comment Number of clusters
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(Labels, n, periodic)$
  \EndProcedure
\end{algorithmic}

@@

Majority of described algorithms contain parallelizable operations such as fast
Fourier transform, cluster labeling and image filtering, hence they are
perfectly suited for execution on GPUs and provide very fast computations (see
execution times below).

\subsection{Directional scanning with line segment}
\label{sec:scan}
All algorithms described above (full correlation maps) are executed on the whole
input image and hence require a lot of RAM memory. The scan approach deals with
memory limitations at the cost of higher execution time (except for $L_2$). With
line-segment scanning we select a list of predefined directions and cut all
one-dimensional slices from the input along those directions. Then for each
direction, as shown in \cref{fig:directions}, we use an appropriate algorithm in
\cref{sec:map} to calculate a correlation function just for that single
slice. For this reason we do not repeat algorithms for $S_2$, $\rho_{ij}$,
$F_{ss}$ and $F_{sv}$.  When all slices are processed we calculate the resulting
directional or averaged CF using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g., image filtering
for edge detection or cluster labeling performed on the whole array, not on
slices.

@@

Functions which calculate full correlation maps are suitable for running on both
CPU and GPU, so we provide execution times for both. The cluster function uses
an algorithm for cluster labeling which has CPU-only implementation (thus,
leaving a room for future improvements). Directional computations are performed
exclusively on CPUs.

@@

While the toolbox of presented package and its API are solid, based on
additional research numerous areas for future improvements of
\code{CorrelationFunctions.jl} are possible:
\begin{enumerate}
\item include novel non-classical correlation functions, e.g., describing
  topology of phases;
\item implementing sub-cube based computations of CFs with a given correlation
  functions with later averaging for larger volume - with such domain
  decomposition one could compute CFs for larger samples with limited GPU
  resources;
\item implementing GPU-based clustering (cluster labeling) and distance
  transform to further speed up computations for $C_2$, chord-length and
  pore-size functions;
\item implementing higher order functions, such as 3-point probabilities or even
  3-point cluster and other more involved CFs;
\item increase number of directions for scan approach or implementing
  calculations within a given solid angle.
\end{enumerate}
We perform steps forward in some directions, whilst others still require
additional research to be made to facilitate their implementation within the
package.
